### 2008-12-19 talk/ecoop2009/rainbow.tex
some improvements, notes

XXX kill the rest?!‽
An interesting effect of virtual structures is that they play nicely with
promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
function for integers, it must first determine that the two arguments
are indeed integer objects.  In the corresponding dispatch logic, we
have added two hints to promote the type of each of the two arguments.
This produces a compiler that has the following behavior: in the general
case, the expression \texttt{a+b} will generate two consecutive run-time
switches followed by the residual code of the proper version of
\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
the intermediate value will contain a compile-time constant as type.
Promoting a compile-time constant is trivial – no run-time code is
generated.  The whole expression \texttt{a+b+c} thus only requires three
switches instead of four.  It is easy to see that even more switches can
be skipped in larger examples; typically, in a tight loop manipulating
only integers, all objects are virtual structures for the compiler and
the residual code is theoretically optimal – all type propagation and
boxing/unboxing occurs at compile-time.

@@

Let us look again at the TLC example.  To ease the reading, figure
\ref{fig:tlc-main} showed a simplified version of TLC's main loop, which did
not include the hints.  The real implementation of the \lstinline{LT} opcode
is shown in figure \ref{fig:tlc-main-hints}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
        elif opcode == LT:
            a, b = stack.pop(), stack.pop()
            hint(a, promote_class=True)
            hint(b, promote_class=True)
            stack.append(IntObj(b.lt(a)))
\end{lstlisting}
\caption{Usage of hints in TLC's main loop}
\label{fig:tlc-main-hints}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
class IntObj(Obj):
    ...
    def lt(self, other): 
        return self.value < other.int_o()
    def int_o(self):
        return self.value
\end{lstlisting}
\caption{Excerpt of the \lstlisting{IntObj} class}
\label{fig:tlc-intobj}
\end{center}
\end{figure}

By promoting the class of \lstlisting{a} and \lstlisting{b}, we tell the JIT
compiler not to generate code until it knows the exact RPython class of both.
Figure \ref{fig:tlc-abs-promotion-1} shows the
code \footnote{\lstinline{switch} is not a legal (R)Python statement, it is
  used here only as a pseudocode example} generated while compiling the usual
\lstlisting{abs} function: note that, compared to figure
\ref{fig:tlc-folded-virtualized}, the code stops just before the calls
\lstlisting{b.lt(a)}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
def interp_eval_abs(args):
    v0 = args[0]
    v1 = IntObj(0)
    a, b = v0, v1
    hint(a, promote_class=True) # no-op
    cls_a = a.__class__
    switch cls_a:
        default: 
            continue_compilation(jitstate, cls_a)
\end{lstlisting}
\caption{Promotion example 1}
\label{fig:tlc-abs-promotion-1}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
def interp_eval_abs(args):
    v0 = args[0]
    v1 = IntObj(0)
    a, b = v0, v1
    hint(a, promote_class=True) # no-op
    cls_a = a.__class__
    switch cls_a:
        IntObj:
            hint(b, promote_class=True)
            v0 = IntObj(b.value < a.value)
            ...
        default: 
            continue_compilation(jitstate, cls_a)
\end{lstlisting}
\caption{Promotion example 2}
\label{fig:tlc-abs-promotion-2}
\end{center}
\end{figure}

>>>>

@@

Let us look again at the TLC example.  To ease the reading, figure
\ref{fig:tlc-main} showed a simplified version of TLC's main loop, which did
not include the hints.  The implementation of the \lstinline{LT} opcode with
hints added is shown in figure \ref{fig:tlc-main-hints}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
def interp_eval(code, pc, args, pool):
    code_len = len(code)
    stack = []
    while pc < code_len:
        opcode = ord(code[pc])
        opcode = hint(opcode, promote_class=True)
        pc += 1

        if opcode == PUSH:
            ...
        elif opcode == LT:
            a, b = stack.pop(), stack.pop()
            hint(a, promote_class=True)
            hint(b, promote_class=True)
            stack.append(IntObj(b.lt(a)))
\end{lstlisting}
\caption{Usage of hints in TLC's main loop}
\label{fig:tlc-main-hints}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
class IntObj(Obj):
    ...
    def lt(self, other): 
        return self.value < other.int_o()
    def sub(self, other):
        return IntObj(self.value - other.int_o())
    def int_o(self):
        return self.value
\end{lstlisting}
\caption{Excerpt of the \lstinline{IntObj} class}
\label{fig:tlc-intobj}
\end{center}
\end{figure}

By promoting the class of \lstinline{a} and \lstinline{b}, we tell the JIT
compiler not to generate code until it knows the exact RPython class of both.
Figure \ref{fig:tlc-abs-promotion-1} shows the
code \footnote{\lstinline{switch} is not a legal (R)Python statement, it is
  used here only as a pseudocode example} generated while compiling the usual
\lstinline{abs} function: note that, compared to figure
\ref{fig:tlc-folded-virtualized}, the code stops just before the calls
\lstinline{b.lt(a)}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
def interp_eval_abs(args):
    v0 = args[0]
    v1 = IntObj(0)
    a, b = v0, v1
    # hint(a, promote_class=True) implemented as follows:
    cls_a = a.__class__
    switch cls_a:
        default: 
            continue_compilation(jitstate, cls_a)
\end{lstlisting}
\caption{Promotion example 1}
\label{fig:tlc-abs-promotion-1}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
def interp_eval_abs(args):
    v0 = args[0]
    v1 = IntObj(0)
    a, b = v0, v1
    # hint(a, promote_class=True) implemented as follows:
    cls_a = a.__class__
    switch cls_a:
        IntObj:
            # hint(b, promote_class=True) needs no code
            v0 = IntObj(b.value < a.value)
            if v0.value:
                return a
            else:
                v0 = IntObj(0)
                v1 = args[0]
                a, b = v0, v1
                v0 = IntObj(b.value - a.value)
                return v0
        default: 
            continue_compilation(jitstate, cls_a)
\end{lstlisting}
\caption{Promotion example 2}
\label{fig:tlc-abs-promotion-2}
\end{center}
\end{figure}
